Modeling site-specific amino-acid preferences deepens phylogenetic estimates of the divergence of viral proteins.

### 2018-04-06 manuscript/manuscript.tex
first pass on introduction comments

Phylogenetics and the concept of a molecular clock are commonly used to estimate the time since the divergence of two lineages, which places evolutionary events on an absolute timescale.
However, in the case of viruses, there is independent evidence that the phylogenetic estimates of deep divergences are too recent. 
For example, integration of filoviruses into their host genomes indicate that the filoviridiae family diverged 7-12 million years ago while the phylogenetic estimate of this event is only $\sim$10,000 years ago~\citep{carroll2013molecular, taylor2014evidence}.  
Similarly, the phylogenetic estimate of when major simian immunodeficiency virus (SIV) groups diverged is almost 100 times more recent than the estimate based on the geographic isolation of SIV host species~\citep{wertheim2009dating, worobey2010island}. 
These examples, along with other similar discrepancies with measles virus~\citep{furuse2010origin, rhazes1848treatise}, coronavirus~\citep{wertheim2013case}, and hepatitis B virus~\citep{fares2002revised}, indicate that phylogenetic methods have a systematic bias toward underestimation of deep branches.

This underestimation is known to occur in part due to inadequate models of purifying selection. 
During evolution protein sites do not sample all amino acids evenly.  
Instead, functional constraint leads to different rates of amino-acid substitution at different sites in the protein and repeated substitutions at a given site~\citep{brown1982mitochondrial}. 
If phylogenetic substitution models fail to account for this constraint, they will underestimate deep branches~\citep{wertheim2011purifying, halpern1998evolutionary}.

Predicting the functional constraints of protein in nature is difficult but laboratory experiments allow for the direct test of a mutation on protein function~\citep{dean2007mechanistic, harms2013evolutionary}. 
Generally in such experiments,  protein variants are generated and then assayed for their functionality relative to wildtype. 
Now, with the the advent of deep mutational scanning~\citep{fowler2014deep}, these experiments can be performed in a high-throughput, comprehensive, and quantitative fashion. 
A deep mutational scan of a protein measured the effect of every amino-acid change on a protein's functionality in lab, which creates a map of the protein's site-specific amino-acid preferences.
These experimental measurements can be incorporated into site-specific phylogenetic models called experimentally informed codon models (ExpCM's) which tend to describe the evolution of a protein better than standard, site-uniform models~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification, haddox2018mapping, lee2018deep, doud2015site}.

Here we use ExpCM's to estimate branch lengths on an influenza hemagglutinin phylogenetic tree and find that these site-specific models estimate longer deep branches. 
We find this extension of branch length is mostly independent from common method of account for purifying selection by rate variation. 
We also found that the ExpCM's estimated similar branch lengths to a site-specific model which infers the amino-acid preferences from the phylogenetic data~\citep{lartillot2004bayesian, rodrigue2014site}.
However, the results from these site-specific models are limited by an effect of purifying selection which is not accounted for in current models - how amino-acid preferences shift over time due to epistasis.

@@

\subsection*{Software versions and computer code}
All code used for the analyses in this paper is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}. \skhcomment{make repo public}
The external computer programs that we used were
\begin{itemize} 
\item \texttt{phydms}~\citep{hilton2017phydms} version 2.2.2 (available at \url{github.com/jbloomlab/phydms}) to fit the ExpCM and GY94 models.
\item \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences.
\item \texttt{PhyloBayes-MPI}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to fit the pbMutSel model. 
\item \texttt{RAxML}~\citep{stamatakis2006raxml} version 8.2.11 (available at \url{https://github.com/stamatak/standard-RAxML} to infer tree topology.
\item We used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization of the results.
\end{itemize}

>>>>

Phylogenetics is commonly used to estimate the time since the divergence of two lineages. 
Under the assumption of a molecular clock, branch lengths on the phylogenetic tree can be transformed into estimates of time. 
When a calibration points exists, these timing estimates of evolutionary events can be placed on an absolute timescale.

However, in the case of viruses, there is independent evidence that the phylogenetic estimates of deep divergences are too recent. 
For example, the integration of filoviruses into their host genomes indicate that the two main clades of filoviridae, ebolovirus and marburg virus, diverged from their common ancestor 7-12 million years ago---but the phylogenetic estimate of this divergence of ebolavirus and margburgvirus is only $\sim$10,000 years ago~\citep{carroll2013molecular, taylor2014evidence}. 
Similarly, the phylogenetic estimate of when major simian immunodeficiency virus (SIV) groups diverged is almost 100 times more recent than the estimate based on the geographic isolation of SIV host species~\citep{wertheim2009dating, worobey2010island}. 
These examples, along with other similar discrepancies with measles virus~\citep{furuse2010origin, rhazes1848treatise}, coronavirus~\citep{wertheim2013case}, and hepatitis B virus~\citep{fares2002revised}, indicate that phylogenetic methods have a systematic bias toward underestimation of deep branches.

This underestimation is known to occur in part due to inadequate models of purifying selection. 
During evolution protein sites do not sample all amino acids evenly.  
Instead, functional constraint leads to different rates of amino-acid substitution at different sites in the protein and repeated substitutions at a given site~\citep{brown1982mitochondrial}. 
If phylogenetic substitution models fail to account for this constraint, they will underestimate deep branches~\citep{wertheim2011purifying, halpern1998evolutionary}. 
Most substitution models account for constraint by trying to statistically infer selection. 
Some models allow the rate of amino-acid substitution to vary among sites according to some statistical distribution~\citep{yang1994maximum,yang2000codon}\skhcomment{models from wertheim2001purifying?}. 
Other models account for this site-specific constraints via site-specific stationary state describing the frequency of every amino acid at every site in the protein~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling, le2008phylogenetic}

A protein's functional constraints can be measured experimentally rather than inferred statistically~\citep{dean2007mechanistic, harms2013evolutionary}. 
Recently, a new experimental approach called deep mutational scanning~\citep{fowler2014deep} has been developed which can measure the site-specific purifying selection quantitatively for all sites in a protein. 
These experimental measurements can be incorporated into site-specific phylogenetic models called experimentally informed codon models (ExpCM's) which tend to describe the evolution of a protein better than standard, site-uniform models~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification, haddox2018mapping, lee2018deep, doud2015site}.

Here we examined how ExpCM's and other forms of modeling purifying selection affects the estimates of deep divergence times of influenza virus hemagglutinin (HA). 
We found that models with site-specific measurements of purifying selection estimates longer deep branches. 
We find this extension of branch length is mostly independent from common method of account for purifying selection by rate variation. 
We also found that the ExpCM's estimated similar branch lengths to a site-specific model which infers the amino-acid preferences from the phylogenetic data~\citep{lartillot2004bayesian, rodrigue2014site}.
However, the results from these site-specific models are limited by an effect of purifying selection which is not accounted for in current models---how amino-acid preferences shift over time due to epistasis.

@@

\subsection*{Software versions and computer code}
All code used for the analyses in this paper is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}. \skhcomment{make repo public}
The external computer programs that we used were
\begin{itemize} 
\item \texttt{phydms}~\citep{hilton2017phydms} version 2.2.2 (available at \url{github.com/jbloomlab/phydms}) to fit the ExpCM and GY94 models.
\item \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences.
\item \texttt{PhyloBayes-MPI}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to fit the pbMutSel model. 
\item \texttt{RAxML}~\citep{stamatakis2006raxml} version 8.2.11 (available at \url{https://github.com/stamatak/standard-RAxML} to infer tree topology.
\item We used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization of the results.
\item \skhcomment{snakemake}
\end{itemize}
